---
title: "A Short Note on Literate Programming"
author: "Jerid Francom"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Short Note on Literate Programming}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Hello world!

**Literate programming** means you can include prose and code into the same document. In R, files with the extension `.Rmd` are R Markdown documents. R Markdown includes the power of Markdown to format text, include graphics, citations, *etc*. and the programming functionality of R (as well as other languages too).

To introduce code into a document we insert a *code chunk* which encloses a section of the document where code will be run and the output returned.

```{r meaningful-chunk-label}
# A comment to describe the code
paste("Hello world!")
```

Code chunks have various *parameters* which allow the user to control various aspects of the behavior of the code. Let's use one right now to get ourselves ready for the rest of the session. `echo=FALSE` turns off the default behavior of showing the code that runs. It will not, however, turn off the output (the `results='hide'` parameter and attribute will do that). 

```{r load-libraries, echo=FALSE}
library(stringr)
```

*Note: no code returned in the rendered document!*

As you work with the other vignettes which cover other topics in R programming in this package, take a look at the source `.Rmd` files. Note the behavior of prose and code in this files to get a sense of how things work!

### More information

To get a basic overview of literate programming in R the [RStudio](https://rmarkdown.rstudio.com) website has documentation that covers the fundamentals of working with the RMarkdown format for various types of output (web documents, PDF documents, Word documents, *etc.*). For a more comprehensive resource consult [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/). 
